---
title: "Malaria Modelling"
author: "Brian Ayanda Manamike"
date: "2025-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Assignment 1 MDS"
author: "Group1"
date: "2025-02-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(epiR)
library(broom)
library(foreign)
library(gtsummary)
library(sf)
library(scales)
library(ggplot2)
library(dplyr)
library(haven)
library(dplyr)
library(tidyr)
library(survey)
library(jtools)
library(remotes)
library(finalfit)
library(kableExtra)
library(ggpubr)
```

```{r}
# Setting our working directory
#setwd("/home/handsom/Group_project_malaria_data_science/Practice-AIMS (1)")
# Load our dataset
#children <- read.csv("dhs-20219-children-data.csv")
children<- read_sav("/home/handsom/data_science/dhs2020/RWKR81SV/RWKR81FL.SAV")
print(head(children))
```

```{r}
#ZZHR62FL <- read_sav("~/Group_project_malaria_data_science/Class DATA/HH/ZZHR62FL.SAV")
setwd("/home/handsom/Group_project_malaria_data_science/dhs2020/RWHR81SV")
ZZHR62FL<- read_sav("/home/handsom/Group_project_malaria_data_science/dhs2020/RWHR81SV/RWHR81FL.SAV")
head(ZZHR62FL)
```
## Here we areb going to group the Child's age variable into three level, **Infants under 2 years**, **Toddlers between 2-3 years** and **Preschoolers between 3-5 years**
```{r}
children <- children %>% 
  mutate(Child_Age=case_when(
    HW1 == 0 | HW1 == 23 ~ "Infants under 2 years",
    HW1 == 24 | HW1 == 35 ~ "Toddlers between 2-3 years",
    HW1 == 36 | HW1 == 59 ~ "Preschoolers between 3-5 years",
    TRUE ~ NA_character_  # Handle unexpected values, assign NA
  ),
  Childs_Age = case_when(
    Child_Age %in% c("Infants under 2 years", "Toddlers between 2-3 years") ~ "Infants and Toddlers under 3 years",
    Child_Age %in% c("Toddlers between 2-3 years", "Preschoolers between 3-5 years") ~ "Toddlers and Preschoolers under 5 years",
    Child_Age == "Infants under 2 years" ~ "Infants under 2 years",
    TRUE ~ "Preschoolers between 3-5 years"
  ))

```

```{r}
table(children$Childs_Age)
```

```{r}
# Create a variable Sex_of_child
children <- children %>%
  mutate(Sex_of_child = case_when(
    B4 == 1 ~ "Male",
    B4 == 2 ~ "Female",
    TRUE ~ NA_character_  # Handle unexpected values, assign NA
  ))
table(children$Sex_of_child)
```

```{r}
# Create a new variable based on the wealth index
children$Wealth_Index <- ifelse(children$V190 %in% c(1, 2), # Poorer or Poorest
                                "poor",
                                ifelse(children$V190 == 3,  # Middle
                                       "Middle",
                                       "Rich"))  # Richer or Richest

# View the updated dataframe
table(children$Wealth_Index)
```

```{r}
# Create a new variable based on the wealth index
children$Educational_level <- ifelse(children$V106==0, # No education
                                "No education",
                                ifelse(children$V106 == 1,  # Primary
                                       "Primary",
                                ifelse(children$V106 == 2,
                                       "Secondary",    # Secondary
                                       "Higher"        #  Higher
                                       )))  

# View the updated dataframe
table(children$Educational_level)
```

 ## This code is creating the target variable `Anti_malaria_taken` by groupng all the ant-amalria drugs taken by individuals.
```{r}
children <- children %>%
  rowwise() %>%
  mutate(
    Anti_malaria_taken = case_when(
      sum(!is.na(c(H37A, H37B, H37C, H37D, H37DA, H37E, H37AA, H37AB, H37F, H37G, H37H))) > 0 &
        sum(c(H37A, H37B, H37C, H37D, H37DA, H37E, H37AA, H37AB, H37F, H37G, H37H), na.rm = TRUE) > 0 ~ 1,
      sum(!is.na(c(H37A, H37B, H37C, H37D, H37DA, H37E, H37AA, H37AB, H37F, H37G, H37H))) == 0 ~ NA,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

# Display the updated data frame
print(children)
```

```{r}
children <- children %>%
  mutate(Anti_Malaria_Taken = case_when(Anti_malaria_taken == 1~ "Yes", 
                                        Anti_malaria_taken == 0~ "No",
                                        TRUE~NA_character_  # Handle unexpected values, assign NA
                                        ))
table(children$Anti_Malaria_Taken)
```

```{r}
# Assuming your data frame is named 'df'

# Create a new variable 'Highest_educational_level'
children$Highest_educational_level <- ifelse(
  children$S108 == 1 | children$S108 == 2,
  "Primary",
  ifelse(children$S108 == 3 | children$S108 == 4,
         "Secondary",
         ifelse(children$S108 == 5, "Higher", NA)
  )
)

# Display the updated data frame
print(children)
```

```{r}
table(children$Highest_educational_level)
```

```{r}
new_children <- children[, c("V001", "V002" ,"Wealth_Index", "Educational_level", "V717", "V481", "V024", "V025","V130", "V394", "V137", "Sex_of_child","Anti_Malaria_Taken", "Highest_educational_level","Childs_Age")]
```

```{r}
colnames(new_children)<- c("V001", "V002" , "Wealth_Index", "Educational_level", "Employement_status", "Health_insurance",  "Region","Type_of_place_of_residence", "Religion", "Visited_health_facility_last_12_month", "Number_of_Under_five", "Sex_of_child","Anti_Malaria_Taken","Highest_educational_level","Childs_Age")
```

```{r}
# Assuming your data frame is named 'responses' and the relevant column is named 'occupation_group'

new_children <- new_children %>%
  mutate(Employment_Status = case_when(
    Employement_status %in% c(1, 2, 3, 4, 5, 6, 7, 8, 9) ~ "Working",           # Categories for Working
    Employement_status == 0 ~ "Not working",                               # Not Working
    Employement_status == 98 ~ "Not working",                              # Don't know                              # Missing
    TRUE ~NA_character_  # Handle unexpected values, assign NA                                             # Handle NA/Not Applicable
  ))

# Display the updated data frame
table(new_children$Employment_Status)
```

## Drop the variable Employement_Status created earlier

```{r}
new_children <- subset(new_children, select = -c(Employement_status))
colnames(new_children)
```

```{r}
# Create a mapping of region numbers to region names
new_children <- new_children %>%
  mutate(Region_Name = case_when(
    Region == 1 ~ "Kigali",
    Region == 2 ~ "South",
    Region == 3 ~ "West",
    Region == 4 ~ "North",
    Region == 5 ~ "East",
    TRUE ~ NA_character_  # Handle unexpected values, assign NA
  ))
table(new_children$Region_Name)
```

# Droping the variable Region we first created

```{r}
new_children <- subset(new_children, select = -c(Region))
colnames(new_children)
```

```{r}
new_children <- new_children %>%
  mutate(Religion_N = case_when(
    Religion %in% c(1, 2, 3, 4, 5, 8, 96) ~ "Belong to a religion",
    Religion == 97 ~ "No religion",
    TRUE ~ NA_character_  # Handle unexpected values, assign NA
  ))
table(new_children$Religion_N)
```

```{r}
new_children <- subset(new_children, select = -c(Religion))
colnames(new_children)
```

```{r}
new_children <- new_children %>%
  mutate(Number_of_Under_five_years = case_when(
    Number_of_Under_five== 0 ~ "No children",
    Number_of_Under_five == 1 ~ "One",
    Number_of_Under_five == 2 ~ "Two", 
    Number_of_Under_five== 3 ~ "Three",
    Number_of_Under_five== 4 ~ "Four",
    Number_of_Under_five== 5 ~ "Five",
    Number_of_Under_five== 6 ~ "Six",
    TRUE ~ NA_character_  # Handle unexpected values, assign NA
  ))
table(new_children$Number_of_Under_five_years)
```

## Here we are dropint the column `Number_of_Under_five` to continue our anlysis with our new column Number_of_Under_five_years

```{r}
new_children <- subset(new_children, select = -c(Number_of_Under_five))
new_children
```

## Categirize the varible `Health_insurance` by creating a new varible `Health_Insurance` which will have $2$ level, `Yes` fir people who are corrently working and `No` for people who are not working.

```{r}
new_children <- new_children %>%
  mutate(Health_Insurance = case_when(
    Health_insurance == 1 ~ "Yes",
    Health_insurance == 0 ~ "No",  # Add a comma here
    TRUE ~ NA_character_  # Handle unexpected values, assign NA
  ))
```

## Droping the column `Health_insurance`.

```{r}
new_children <- subset(new_children, select = -c(Health_insurance))
new_children
```

## Changing the level of the categorical variable `Visited_health_facility_last_12_month` into `Yes` and `No`.

```{r}
new_children <- new_children %>%
  mutate(Visited_health_Facility_Last_12_Month = case_when(
    Visited_health_facility_last_12_month == 1 ~ "Yes",
    Visited_health_facility_last_12_month == 0 ~ "No",  # Add a comma here
    TRUE ~ NA_character_  # Handle unexpected values, assign NA
  ))
```

## Droping the column `Visited_health_facility_last_12_month` which was replace by `Visited_health_Facility_Last_12_Month`

```{r}
new_children <- subset(new_children, select = -c(Visited_health_facility_last_12_month))
new_children
```

```{r}
# Create new variable Place_of_residence
new_children <- new_children %>%
  mutate(Place_of_residence = ifelse(Type_of_place_of_residence == 1, "Urban", "Rural"))
table(new_children$Place_of_residence)
```

```{r}
new_children <- subset(new_children, select = -c(Type_of_place_of_residence))
colnames(new_children)
```

```{r}
SS3 <- data.frame(V1=ZZHR62FL$HV001,
                 V2=ZZHR62FL$HV002,
                 SB1=ZZHR62FL$`SBSEL$01`,
                 SB2=ZZHR62FL$`SBSEL$02`,
                 SB3=ZZHR62FL$`SBSEL$03`,
                 SB4=ZZHR62FL$`SBSEL$04`,
                 SB5=ZZHR62FL$`SBSEL$05`,
                 SB6=ZZHR62FL$`SBSEL$06`,
                 SB7=ZZHR62FL$`SBSEL$07`,
                 SB8=ZZHR62FL$`SBSEL$08`,
                 SB9=ZZHR62FL$`SBSEL$09`,
                 SB10=ZZHR62FL$`SBSEL$10`,
                 SB11=ZZHR62FL$`SBSEL$11`,
                 SB12=ZZHR62FL$`SBSEL$12`,
                 SB13=ZZHR62FL$`SBSEL$13`,
                 SB14=ZZHR62FL$`SBSEL$14`,
                 SB15=ZZHR62FL$`SBSEL$15`,
                 SB16=ZZHR62FL$`SBSEL$16`)
print(head(SS3))
```

```{r}
View(SS3)
```

```{r}
SS4 <- data.frame(V1= children$V001,
                  V2= children$V002,
                    B16= children$B16)

```


## Setting the merging in other to count chilgren that vae malaria
```{r}
SS3$maching <- paste(SS3$V1, SS3$V2, sep = "_")
SS4$maching <- paste(SS4$V1, SS4$V2, sep = "_")
merge2 <- merge(SS3, SS4, by="maching")

```

```{r}
# Convert SB columns to character in advance
merge2 <- merge2 %>%
  mutate(across(SB1:SB16, as.character))

# Use rowwise and mutate
merge2 <- merge2 %>%
  rowwise() %>%
  mutate(Result_test = if (!is.na(B16) && B16 > 0) {
      if (any(c_across(SB1:SB16) == 2, na.rm = TRUE)) {
        "Positive"
      } else {
        "Negative"
      }
    } else {
      NA  # Here, NA or any other label if needed
    }) %>%
  ungroup()  # Ungroup after row-wise operation
```

```{r}
table(merge2$Result_test)
```

## Final merging

```{r}
new_children$maching <- paste(new_children$V001, new_children$V002, sep = "_")
Filnal_merge <- merge(new_children, merge2, by="maching")
View(Filnal_merge)
```

```{r}
Final_dataset<- Filnal_merge[, c("Anti_Malaria_Taken","Wealth_Index","Result_test" ,"Sex_of_child","Highest_educational_level","Employment_Status","Region_Name" ,"Religion_N"                 ,"Number_of_Under_five_years" ,"Health_Insurance"                     
 ,"Visited_health_Facility_Last_12_Month" ,"Place_of_residence","Childs_Age")]

```

```{r}

```

```{r}
## Removing people that have a negative test
Final_dataset_cleaned <- Final_dataset[Final_dataset$Result_test != "Negative", ]

```

```{r}
Final_dataset_cleaned <- Final_dataset_cleaned[!is.na(Final_dataset_cleaned$Anti_Malaria_Taken), ]

```

```{r}
colnames(Final_dataset_cleaned)
```

## Descriptive analysis Socio-Economic factors

```{r}
v_indep <- c("Wealth_Index","Highest_educational_level", "Health_Insurance", "Employment_Status","Childs_Age"  , "Number_of_Under_five_years","Sex_of_child","Visited_health_Facility_Last_12_Month")
v_dep <- "Anti_Malaria_Taken"  
Final_dataset_cleaned |> summary_factorlist(v_dep, v_indep, p=TRUE, p_cont_para = "t.test", p_cat = "chisq", add_dependent_label = FALSE, cont = "mean", total_col = FALSE, add_col_totals = TRUE, add_row_totals = TRUE, na_include = TRUE, na_to_p = FALSE, na_to_prop = FALSE, include_col_totals_percent = FALSE) -> Final_dataset_summary

#xlsx:: write.xlsx(Merge_file_summary, "Merge_file_summary.xlsx", sheetName="1.2b", append=FALSE, row.names=FALSE)

Final_dataset_summary |>
  kable(caption = "") |>
  kable_styling() |>
  kable_classic(lightable_options = c("striped", "hover"))
```

```{r}
Final_dataset_summary <- Final_dataset_cleaned %>%
  tbl_summary(
    by = "Anti_Malaria_Taken", 
    statistic = all_categorical() ~ "{n} ({p}%)"
  )
Final_dataset_summary
```

## Multivariate analysis for socio-economic factor

```{r}
Final_dataset_cleaned$Anti_Malaria_Taken_binary <- as.integer(Final_dataset_cleaned$Anti_Malaria_Taken =="Yes")
# If missing values are present, you might decide to remove them
# However, removing rows with missing values can lead to biased results
# Here's how you could do it, but consider imputation or other strategies
#Final_dataset_complete <- na.omit(Final_dataset)

modelEC1 <- glm (Anti_Malaria_Taken_binary~ Wealth_Index + Highest_educational_level  + Number_of_Under_five_years + Employment_Status + Childs_Age + Visited_health_Facility_Last_12_Month + Sex_of_child, family=binomial(), data=Final_dataset_cleaned)

summary(modelEC1)

```

```{r}
modelEC1 |> 
  tbl_regression(exponentiate = TRUE,
                 label=list(Wealth_Index~ "Wealth Index",
                            Highest_educational_level ~ "Highest Educational level", 
                           Employment_Status ~ "Currently Insurence",
                           Childs_Age ~ "Age of Children",
                          Visited_health_Facility_Last_12_Month ~ "Visited Health Facility Last 12 Month",
                          Sex_of_child~ "Sex of child"))  |> 
  bold_labels() |> 
  bold_p()
```

## Demographic and Geographic factors

```{r}
v_indep <- c("Region_Name","Religion_N", "Childs_Age", "Place_of_residence" ,"Sex_of_child" )
v_dep <- "Anti_Malaria_Taken" 
Final_dataset |> summary_factorlist(v_dep, v_indep, p=TRUE, p_cat = "chisq", add_dependent_label = FALSE, cont = "mean", total_col = FALSE, add_col_totals = TRUE, add_row_totals = TRUE, na_include = TRUE, na_to_p = FALSE, na_to_prop = FALSE, include_col_totals_percent = FALSE) -> Merge_file_summary

#xlsx:: write.xlsx(Merge_file_summary, "Merge_file_summary.xlsx", sheetName="1.2b", append=FALSE, row.names=FALSE)

Merge_file_summary |>
  kable(caption = "") |>
  kable_styling() |>
  kable_classic(lightable_options = c("striped", "hover"))
```

## Mutivariate analysis of the Geographic and Demographic factors

```{r}
modelEC2 <- glm (Anti_Malaria_Taken_binary~ Region_Name + Childs_Age   + Number_of_Under_five_years + Place_of_residence + Sex_of_child + Religion_N, family=binomial(), data=Final_dataset_cleaned)

summary(modelEC2)

```
## When we perfom the `glm` on the Demographic and Geographic factors, we can see that the p-value of the `South` and the `West` region is less than $0.05$ meaning that these two region influence a lot on the disparities of anti-malaria drugs in Rwanda but we were expecting also to see some varibles like `Number of children in the household` because if if a household has too many children, if there don't have anough money it will be more likly that children will not have acces on the anti-malaria drug espacially here in Africa where there is malaria in almost all the countrie and sometime, all the children in the household can be infected.


```{r}
modelEC2 |> 
  tbl_regression(exponentiate = TRUE,
                 label=list(Region_Name~ "Region names",
                             Number_of_Under_five_years ~ "Number of child under five years in the household", 
                           Childs_Age ~ "Age of Children",
                          Religion_N ~ "Religion",
                          Sex_of_child~ "Sex of child"))  |> 
  bold_labels() |> 
  bold_p()
```



```{r}
Final_dataset_cleaned |>  group_by(Region_Name, Anti_Malaria_Taken) |> 
  summarize(n=n()) |> 
  mutate(p=n/sum(n),
pct=scales:: percent(p, accuracy = 0.1)) -> e1_2d

ggplot(e1_2d, aes(x=Region_Name, y=p, fill = Anti_Malaria_Taken))+
  geom_bar(stat= "identity", position = "fill")+
  scale_y_continuous(breaks = seq(0,1,.2), labels = percent)+
  geom_text(aes(label = pct), size = 3, position = position_stack(vjust = 0.5))+
  scale_fill_brewer(palette = "Set2")+
  labs(y="Percent", fill="Anti-malaria Drug", x="Region")+
  theme_pubr()
```




